Identifying Effective Motivational Interviewing Communication Sequences Using Automated Pattern Analysis

### 2018-06-29 jhir/latex-manuscript/manuscript.tex
submitted to JHIR as minor revision

\usepackage{rotating}
\usepackage{tabularx}
\usepackage{longtable,tabu}
\usepackage[flushleft]{threeparttable}
\usepackage{caption}
\captionsetup{justification=raggedright,singlelinecheck=off,labelfont=bf}
\usepackage{color, colortbl}
\usepackage{soul}
\definecolor{lightgraycol}{gray}{0.95}

@@

\title{\hl{Identifying Effective Motivational Interviewing Communication Sequences Using Automated Pattern Analysis}
}

@@

\begin{abstract}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change. Although there is strong empirical evidence linking ``MI-consistent'' counselor behaviors and patient motivational statements (i.e., ``change talk''), the specific counselor communication behaviors effective for eliciting patient change talk vary by treatment context and, thus, are a subject of ongoing research. \hl{An integral part of this research is the sequential analysis of pre-coded MI transcripts.} In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice. We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers. Transcripts had been previously annotated with patient-counselor behavior codes using a specialized codebook. Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify counselor communication strategies that are effective at eliciting patients' motivational statements to guide clinical practice.

@@

Sequential analysis is an analytic approach to examine temporally ordered sequences of events or observations~\cite{bakeman1997observing, bakeman2011sequential}. Moyers and Martin~\cite{moyers2006therapist} applied sequential analysis in a study of adults in treatment for alcohol abuse and \hl{found out that change talk} was significantly more likely after MICO and sustain talk more likely after MIIN. A follow-up study with the same population \hl{found that change talk} was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk~\cite{moyers2009session}. Surprisingly, MIIN was unrelated to sustain talk, but decreased the likelihood of change talk. Gaume et al.~\cite{gaume2008counsellor} used sequential analysis to study communication patterns during brief motivational interviewing for hazardous alcohol consumption with young adults conscripted into military service. They found \hl{that} MICO led to both change talk and sustain talk but the MIIN-to-sustain talk pattern was not observed. A second study with the same population confirmed that MICO leads to significantly more change talk and sustain talk~\cite{gaume2010counselor}. In this sample, MIIN led to greater sustain talk, but was unrelated to change talk. Further analyses revealed that reflections were the only MICO behavior linked to increased change talk; reflections and other MICO behaviors, excluding questions, were related to increased sustain talk. Glynn and colleagues~\cite{glynn2014change} linked reflections of change talk to the elicitation of change talk and reflections of sustain talk to the elicitation of sustain talk among incarcerated adolescents with high rates of alcohol and marijuana use. In a study of adolescents engaged in weight loss treatment, Carcone et al.~\cite{carcone2013provider} used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk and statements emphasizing decision-making autonomy. A parallel study of the adolescents' caregivers~\cite{jacques2016building} drew a similar conclusion \hl{that} asking questions phrased to elicit change talk, reflections of change talk and \hl{autonomy-supportive} statements \hl{were} the counselor behaviors, \hl{which led} to the elicitation of change talk. Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors\hl{, however, }led to change talk in some \hl{treatment contexts, but not others,} suggesting a need for additional research to understand the \hl{treatment contexts, in which various} MICO strategies are effective. \hl{The current study contributes to existing knowledge by examining African American adolescents in weight loss treatment.}

The sequential analysis procedure used in the above MI process studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} is based on the first-order Markov Chain model~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}. The Markov Chain model is a discrete-time stochastic process built on the assumption that the state of a system or condition changes over time and only depends on the previous event. Hence, \hl{Markov Chain models} have two main drawbacks. The first is their inability to preserve the long-range dependencies between observations in a sequence. However, in MI, an observed behavior can be influenced by any of the preceding behaviors. The second drawback is their inability to consider similarities between behavior codes and, consequently, their inability to identify multiple similar behaviors that lead to the same outcome. Thus, first-order Markov models may be insufficient to fully understand the associations between behaviors in patient-counselor communication sequences. Therefore, there is a need for more powerful computational methods, which consider clusters of similar behavior codes and \hl{long-range} dependencies between behaviors, to identify causal relationships. The goal of the current research is to test the applicability \hl{of data mining and machine learning} methods to identify effective patterns of patient-counselor communication.

Several prior works have reported the results of adopting machine learning methods, such as topic models~\cite{kotov2014towards, gaut2017content, imel2015computational, kotov2015interpretable, atkins2014scaling}, classification methods~\cite{hasan2016study, can2016sounds, perez2017predicting, perez2017understanding} and neural networks~\cite{hasan2016study, tanana2015recursive, tanana2016comparison} to the tasks of annotating MI transcripts and assessing intervention fidelity. Perez-Rosas et al.~\cite{perez2017predicting} developed a natural language processing system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational \hl{interviews. Their} system employed \hl{an SVM} classifier based on n-grams (contiguous sequences of words of a specified length), syntactic and \hl{semantic features}. In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes~\cite{hasan2016study}. The SVM model with the aforementioned features achieved 75\% accuracy for automated annotation of MI transcripts with 17 behavior codes, which is comparable to human coders. In a follow-up study~\cite{hasan2018predicting}, we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk). RNN achieved 87\% accuracy, 17\% greater than MMs (70\%) in predicting the success of motivational interviews. \hl{The current work builds on past work performed on automation of annotation.} We examine the efficacy of Hidden Markov Models (HMMs) and frequent pattern mining in identifying the counselor communication strategies antecedent to patient change talk.

HMMs are widely used for the analysis of sequence data due to their ability to model \hl{long-range dependencies} between clusters of discrete observations in a sequence. The HMM associates each observation in a sequence with a ``hidden'' state, which corresponds to a distribution over all distinct observations in a sequence (i.e., probabilities associated with each observation, when HMM is in this hidden state), such that each ``hidden state'' corresponds to a different distribution. Sequences of observations are modeled as transitions between different hidden states and sampling observations from distributions corresponding to each hidden state. HMMs were originally proposed for speech recognition~\cite{rabiner1989tutorial}, in which the states were used to represent all English language sounds. In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling~\cite{uuguz2007biomedical, amrhein2003client}. For example, \hl{an HMM-based} classifier was \hl{applied to Doppler} ultrasound imaging data to extract features \hl{from} the images that were then used to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}. In another study, HMM was used to capture important
characteristics of protein families~\cite{amrhein2003client}. In the application of HMM to patient-counselor communication, hidden states and the sets of related behavior codes associated with the hidden states \hl{may correspond to patients' underlying} motivational state during a patient-counselor encounter. Although the MI literature has established patient change talk and commitment language (a special class of change talk where patients express their intentions, plans and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patients' behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk. Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that are likely to lead to patient change talk and commitment language.

Frequent pattern mining~\cite{aggarwal2014frequent} is a class of data mining methods to identify sets of items (or observations, referred to as itemsets) which frequently appear together. Agrawal and Srikant~\cite{agrawal1994fast} first introduced frequent pattern mining with the Apriori algorithm, which they developed to identify patterns in customer purchases. Since its introduction, frequent pattern mining has been applied to several other domains, including health informatics~\cite{abdullah2008analysis, olukunle2002fast, bethel2006mining, wright2015use}, medical imaging~\cite{olukunle2002fast}, chemical and biological analysis~\cite{deshpande2005frequent, leach2007introduction, zhang2008discovering}, web mining~\cite{srivastava2000web} and outlier analysis~\cite{aggarwal2015outlier}. \hl{However, no published} research has yet examined the utility of frequent pattern mining for studying patient-counselor communication. A major challenge in applying frequent pattern mining methods to patient-counselor communication sequences is the large number of resulting patterns, which include redundant patterns. To address this problem, we utilized the closed frequent itemset mining method~\cite{pasquier1999discovering}, which produces fewer patterns in \hl{a more} compact form that are easier to interpret. In this study, we leveraged FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that frequently lead to patient change talk. \hl{FPClose is a state-of-the-art closed} frequent itemset mining algorithm, which has \hl{demonstrated good performance} in terms of running time and memory consumption.

In this paper, we focus on computational methods to facilitate the sequential analysis \hl{of pre-coded MI transcripts} to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions. Analysis of these patterns will provide empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk. This knowledge will inform MI theory by providing additional evidence to support MI's technical hypothesis. It will also inform clinical practice by facilitating the use of more effective and tailored counselor communication. This paper is the first empirical evaluation of the effectiveness of closed frequent pattern mining to analyze patient-counselor communication sequences during MI sessions. Bertholet et al.~\cite{bertholet2010change} used HMM to identify hidden states in a brief motivational intervention. Namely, they limited their HMM model to three hidden states which they characterized as ``towards change'', ``away from change'' and ``non-determined''. These states were then used as predictors in a bivariate regression model to predict drinking outcomes 12 months post-intervention. In the current study, we identified the optimal number of hidden states using HMM modeling of successful and unsuccessful sequences of patient-counselor communication. The goal of this study to evaluate the utility of using HMM and frequent pattern mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions. These two approaches offer the following advantages over the first-order \hl{Markov Chain-based} methods most typically used in MI research. First-order \hl{Markov Chain} models identify the likely transitions between individual behaviors. In contrast, HMM summarizes transitions between clusters of related behavior codes (i.e., hidden states) allowing the identification of clusters of behaviors antecedent to change talk in successful patient-counselor communications and sustain talk in unsuccessful patient-counselor communications. Frequent pattern mining can identify patterns involving long-range dependencies between patient and counselor behaviors. Accounting for such long-range dependencies is important, since human behaviors, such as patient-counselor communications during MI sessions, are informed by all the antecedent behaviors and not just the immediately preceding behavior.

@@

\subsection{Dataset}
\label{subsec:dataset}
The \hl{dataset} for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss. Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor. These encounters \hl{had been previously} annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider} \hl{excluding conversations correspond to greetings, farewell and interview setups such as table and camera settings.} The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions. The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational} and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}. A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$). The experimental dataset consists of 7,192 patient, caregiver and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.

@@

\rowcolor{lightgraycol}
\noalign{\smallskip}
\multicolumn{4}{l}{\textbf{Patient}} \\
\noalign{\smallskip}
CT & Change Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans and action steps) changing their behavior & ``I will try to buy less junk food.''\\
ST & Sustain Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans and action steps) to maintain the status quo or not change their behavior  & ``I didn't get to the gym this week.''\\
\hl{HUPW} & \hl{High uptake weight} & \hl{A turn that does develop the topic of the conversation. High Uptake statements include: weight related statements about actions of commitment, change talk and ambivalence that occurred in the past, patient questions to the counselor and session interruptions by persons who are not an active part of the treatment session.} & \hl{``Support is always good.  You know that's a key factor.  Mm-hmm.''}\\
\hl{HUPO} & \hl{High uptake other} & \hl{An utterance that develops the topic of the conversation but is about non-target behaviors or interruptions} & \hl{``Yeah because the mentor comes and they take off and they go someplace for a little while.''}\\
LUP & Low uptake  & An utterance that does not develop the topic of conversation but still allows it to continue & ``Mm-hmm.  Right.''\\
\hline
\end{longtable}
\end{small}

@@

\begin{itemize}
\item $M$ is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table~\ref{tab:codebook})
\item $T$ is an \hl{$N \times N$} state transition probability matrix, in which $t_{ij}$ is the probability of HMM transitioning from state $i$ to state $j$  
\item $E$ is an \hl{$N \times M$} emission probability matrix, in which $e_{jk}$ is the probability of observing symbol $k$, when HMM is in state $j$ 
\item \hl{$\pi$} is the initial state distribution vector where $\pi_i$ is the probability of \hl{the $i^{th}$ state to be the first state}
\end{itemize}

@@

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/bic-vs-hidden-states.eps}
\caption{\hl{Bayesian information criterion (BIC) of HMM models of successful (left) and unsuccessful (right) interviews by varying the number of hidden states}}
\label{fig:bic}
\end{figure*}

\subsubsection{Frequent Pattern Mining}
\label{subsubsec:frequentPatternMining}
We applied frequent pattern mining to identify frequently occurring \hl{patterns of patient-counselor} behavior codes in successful and unsuccessful communication sequences. Behavior codes in these patterns may be separated by one or more other codes. For this purpose, we utilized FPClose~\cite{grahne2005fast}, an efficient state-of-the-art \hl{closed} frequent pattern mining algorithm implemented in SPMF~\cite{fournier2014spmf, fournier2016spmf}, to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences. SPMF is an open-source library providing more than 150 data mining algorithms. Popular \hl{non-closed} frequent pattern mining algorithms include Apriori~\cite{agrawal1994fast} and FP-Growth~\cite{han2000mining}. A \textit{frequent pattern} is defined as a \hl{pattern of observations}, which appears in a given set of sequences more often than a user-specified threshold called the \textit{minimum support count}. For example, \{A\}, \{C\}, \{D\} and \{C, D\} are frequent patterns in the example set of sequences in Figure~\ref{fig:fpm}, since these patterns appear at least 2 times, which is the minimum support count in this example. In this work, we identified and analyzed \textit{closed frequent patterns} among all sequences of behavior codes in successful and unsuccessful communication sequences. A \textit{frequent pattern} is \textit{closed} if none of its supersets have the same support count~\cite{pasquier1999discovering}, where a set X is a superset of another set Y, if X contains all the elements of the set Y. For example, the itemsets \{A\}, \{C\}, \{D\} and \{C, D\} in Figure~\ref{fig:fpm} are not closed frequent \hl{patterns since} their supersets \{A, B\}, \{B, C\}, \{B, D\} and \{B, C, D\} have the same support count. Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\} and \{B, C, D\} are closed frequent \hl{patterns since} none of their supersets have the same support count. On the other hand, itemsets \{A, C\}, \{A, B, C\}, \{A, B, D\}, \{A, C, D\} and \{A, B, C, D\} have support counts of 1, 0, 1, 0 and 1, respectively, which is less than the minimum support count and thus are identified as non-frequent itemsets. Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed \hl{prior work}~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences. For each pattern, the statistical significance \hl{of the difference} between pattern occurrence in successful and unsuccessful communications was reported by computing \hl{p-value of Pearson's} chi-square test.

>>>>

\usepackage{rotating}
\usepackage{tabularx}
\usepackage{longtable,tabu}
\usepackage[flushleft]{threeparttable}
\usepackage{caption}
\captionsetup{justification=raggedright,singlelinecheck=off,labelfont=bf}
\usepackage{color, colortbl}
\definecolor{lightgraycol}{gray}{0.95}

@@

\title{Identifying Effective Motivational Interviewing Communication Sequences Using Automated Pattern Analysis
}

@@

\begin{abstract}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change. Although there is strong empirical evidence linking ``MI-consistent'' counselor behaviors and patient motivational statements (i.e., ``change talk''), the specific counselor communication behaviors effective for eliciting patient change talk vary by treatment context and, thus, are a subject of ongoing research. An integral part of this research is the sequential analysis of pre-coded MI transcripts. In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice. We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers. Transcripts had been previously annotated with patient-counselor behavior codes using a specialized codebook. Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify counselor communication strategies that are effective at eliciting patients' motivational statements to guide clinical practice.

@@

Sequential analysis is an analytic approach to examine temporally ordered sequences of events or observations~\cite{bakeman1997observing, bakeman2011sequential}. Moyers and Martin~\cite{moyers2006therapist} applied sequential analysis in a study of adults in treatment for alcohol abuse and found that change talk was significantly more likely after MICO and sustain talk more likely after MIIN. A follow-up study with the same population found that change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk~\cite{moyers2009session}. Surprisingly, MIIN was unrelated to sustain talk, but decreased the likelihood of change talk. Gaume et al.~\cite{gaume2008counsellor} used sequential analysis to study communication patterns during brief motivational interviewing for hazardous alcohol consumption with young adults conscripted into military service. They found that MICO led to both change talk and sustain talk but the MIIN-to-sustain talk pattern was not observed. A second study with the same population confirmed that MICO leads to significantly more change talk and sustain talk~\cite{gaume2010counselor}. In this sample, MIIN led to greater sustain talk, but was unrelated to change talk. Further analyses revealed that reflections were the only MICO behavior linked to increased change talk; reflections and other MICO behaviors, excluding questions, were related to increased sustain talk. Glynn and colleagues~\cite{glynn2014change} linked reflections of change talk to the elicitation of change talk and reflections of sustain talk to the elicitation of sustain talk among incarcerated adolescents with high rates of alcohol and marijuana use. In a study of adolescents engaged in weight loss treatment, Carcone et al.~\cite{carcone2013provider} used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk and statements emphasizing decision-making autonomy. A parallel study of the adolescents' caregivers~\cite{jacques2016building} drew a similar conclusion that asking questions phrased to elicit change talk, reflections of change talk and autonomy-supportive statements were the counselor behaviors, which led to the elicitation of change talk. Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors, however, led to change talk in some treatment contexts, but not others, suggesting a need for additional research to understand the treatment contexts, in which various MICO strategies are effective. The current study contributes to existing knowledge by examining African American adolescents in weight loss treatment.

The sequential analysis procedure used in the above MI process studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} is based on the first-order Markov Chain model~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}. The Markov Chain model is a discrete-time stochastic process built on the assumption that the state of a system or condition changes over time and only depends on the previous event. Hence, Markov Chain models have two main drawbacks. The first is their inability to preserve the long-range dependencies between observations in a sequence. However, in MI, an observed behavior can be influenced by any of the preceding behaviors. The second drawback is their inability to consider similarities between behavior codes and, consequently, their inability to identify multiple similar behaviors that lead to the same outcome. Thus, first-order Markov models may be insufficient to fully understand the associations between behaviors in patient-counselor communication sequences. Therefore, there is a need for more powerful computational methods, which consider clusters of similar behavior codes and long-range dependencies between behaviors, to identify causal relationships. The goal of the current research is to test the applicability of data mining and machine learning methods to identify effective patterns of patient-counselor communication.

Several prior works have reported the results of adopting machine learning methods, such as topic models~\cite{kotov2014towards, gaut2017content, imel2015computational, kotov2015interpretable, atkins2014scaling}, classification methods~\cite{hasan2016study, can2016sounds, perez2017predicting, perez2017understanding} and neural networks~\cite{hasan2016study, tanana2015recursive, tanana2016comparison} to the tasks of annotating MI transcripts and assessing intervention fidelity. Perez-Rosas et al.~\cite{perez2017predicting} developed a natural language processing system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews. Their system employed an SVM classifier based on n-grams (contiguous sequences of words of a specified length), syntactic and semantic features. In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes~\cite{hasan2016study}. The SVM model with the aforementioned features achieved 75\% accuracy for automated annotation of MI transcripts with 17 behavior codes, which is comparable to human coders. In a follow-up study~\cite{hasan2018predicting}, we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk). RNN achieved 87\% accuracy, 17\% greater than MMs (70\%) in predicting the success of motivational interviews. The current work builds on past work performed on automation of annotation. We examine the efficacy of Hidden Markov Models (HMMs) and frequent pattern mining in identifying the counselor communication strategies antecedent to patient change talk.

HMMs are widely used for the analysis of sequence data due to their ability to model long-range dependencies between clusters of discrete observations in a sequence. The HMM associates each observation in a sequence with a ``hidden'' state, which corresponds to a distribution over all distinct observations in a sequence (i.e., probabilities associated with each observation, when HMM is in this hidden state), such that each ``hidden state'' corresponds to a different distribution. Sequences of observations are modeled as transitions between different hidden states and sampling observations from distributions corresponding to each hidden state. HMMs were originally proposed for speech recognition~\cite{rabiner1989tutorial}, in which the states were used to represent all English language sounds. In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling~\cite{uuguz2007biomedical, amrhein2003client}. For example, an HMM-based classifier was applied to Doppler ultrasound imaging data to extract features from the images that were then used to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}. In another study, HMM was used to capture important
characteristics of protein families~\cite{amrhein2003client}. In the application of HMM to patient-counselor communication, hidden states and the sets of related behavior codes associated with the hidden states may correspond to patients' underlying motivational state during a patient-counselor encounter. Although the MI literature has established patient change talk and commitment language (a special class of change talk where patients express their intentions, plans and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patients' behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk. Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that are likely to lead to patient change talk and commitment language.

Frequent pattern mining~\cite{aggarwal2014frequent} is a class of data mining methods to identify sets of items (or observations, referred to as itemsets) which frequently appear together. Agrawal and Srikant~\cite{agrawal1994fast} first introduced frequent pattern mining with the Apriori algorithm, which they developed to identify patterns in customer purchases. Since its introduction, frequent pattern mining has been applied to several other domains, including health informatics~\cite{abdullah2008analysis, olukunle2002fast, bethel2006mining, wright2015use}, medical imaging~\cite{olukunle2002fast}, chemical and biological analysis~\cite{deshpande2005frequent, leach2007introduction, zhang2008discovering}, web mining~\cite{srivastava2000web} and outlier analysis~\cite{aggarwal2015outlier}. However, no published research has yet examined the utility of frequent pattern mining for studying patient-counselor communication. A major challenge in applying frequent pattern mining methods to patient-counselor communication sequences is the large number of resulting patterns, which include redundant patterns. To address this problem, we utilized the closed frequent itemset mining method~\cite{pasquier1999discovering}, which produces fewer patterns in a more compact form that are easier to interpret. In this study, we leveraged FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that frequently lead to patient change talk. FPClose is a state-of-the-art closed frequent itemset mining algorithm, which has demonstrated good performance in terms of running time and memory consumption.

In this paper, we focus on computational methods to facilitate the sequential analysis of pre-coded MI transcripts to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions. Analysis of these patterns will provide empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk. This knowledge will inform MI theory by providing additional evidence to support MI's technical hypothesis. It will also inform clinical practice by facilitating the use of more effective and tailored counselor communication. This paper is the first empirical evaluation of the effectiveness of closed frequent pattern mining to analyze patient-counselor communication sequences during MI sessions. Bertholet et al.~\cite{bertholet2010change} used HMM to identify hidden states in a brief motivational intervention. Namely, they limited their HMM model to three hidden states which they characterized as ``towards change'', ``away from change'' and ``non-determined''. These states were then used as predictors in a bivariate regression model to predict drinking outcomes 12 months post-intervention. In the current study, we identified the optimal number of hidden states using HMM modeling of successful and unsuccessful sequences of patient-counselor communication. The goal of this study to evaluate the utility of using HMM and frequent pattern mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions. These two approaches offer the following advantages over the first-order Markov Chain-based methods most typically used in MI research. First-order Markov Chain models identify the likely transitions between individual behaviors. In contrast, HMM summarizes transitions between clusters of related behavior codes (i.e., hidden states) allowing the identification of clusters of behaviors antecedent to change talk in successful patient-counselor communications and sustain talk in unsuccessful patient-counselor communications. Frequent pattern mining can identify patterns involving long-range dependencies between patient and counselor behaviors. Accounting for such long-range dependencies is important, since human behaviors, such as patient-counselor communications during MI sessions, are informed by all the antecedent behaviors and not just the immediately preceding behavior.

@@

\subsection{Dataset}
\label{subsec:dataset}
The dataset for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss. Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor. These encounters had been previously annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider} excluding conversations that correspond to greetings, farewell and interview setups such as table and camera settings. The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions. The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational} and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}. A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$). The experimental dataset consists of 7,192 patient, caregiver and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.

@@

\rowcolor{lightgraycol}
\noalign{\smallskip}
\multicolumn{4}{l}{\textbf{Patient}} \\
\noalign{\smallskip}
CT & Change Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans and action steps) changing their behavior & ``I will try to buy less junk food.''\\
ST & Sustain Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans and action steps) to maintain the status quo or not change their behavior  & ``I didn't get to the gym this week.''\\
HUPW & High uptake weight & A turn that does develop the topic of the conversation. High Uptake statements include: weight related statements about actions of commitment, change talk and ambivalence that occurred in the past, patient questions to the counselor and session interruptions by persons who are not an active part of the treatment session. & ``Support is always good.  You know that's a key factor.  Mm-hmm.''\\
HUPO & High uptake other & An utterance that develops the topic of the conversation but is about non-target behaviors or interruptions & ``Yeah because the mentor comes and they take off and they go someplace for a little while.''\\
LUP & Low uptake  & An utterance that does not develop the topic of conversation but still allows it to continue & ``Mm-hmm.  Right.''\\
\hline
\end{longtable}
\end{small}

@@

\begin{itemize}
\item $M$ is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table~\ref{tab:codebook})
\item $T$ is an $N \times N$ state transition probability matrix, in which $t_{ij}$ is the probability of HMM transitioning from state $i$ to state $j$  
\item $E$ is an $N \times M$ emission probability matrix, in which $e_{jk}$ is the probability of observing symbol $k$, when HMM is in state $j$ 
\item $\pi$ is the initial state distribution vector where $\pi_i$ is the probability of the $i^{th}$ state to be the first state
\end{itemize}

@@

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/bic-vs-hidden-states.eps}
\caption{Bayesian information criterion (BIC) of HMM models of successful (left) and unsuccessful (right) interviews by varying the number of hidden states}
\label{fig:bic}
\end{figure*}

\subsubsection{Frequent Pattern Mining}
\label{subsubsec:frequentPatternMining}
We applied frequent pattern mining to identify frequently occurring patterns of patient-counselor behavior codes in successful and unsuccessful communication sequences. Behavior codes in these patterns may be separated by one or more other codes. For this purpose, we utilized FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining algorithm implemented in SPMF~\cite{fournier2014spmf, fournier2016spmf}, to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences. SPMF is an open-source library providing more than 150 data mining algorithms. Popular non-closed frequent pattern mining algorithms include Apriori~\cite{agrawal1994fast} and FP-Growth~\cite{han2000mining}. A \textit{frequent pattern} is defined as a pattern of observations, which appears in a given set of sequences more often than a user-specified threshold called the \textit{minimum support count}. For example, \{A\}, \{C\}, \{D\} and \{C, D\} are frequent patterns in the example set of sequences in Figure~\ref{fig:fpm}, since these patterns appear at least 2 times, which is the minimum support count in this example. In this work, we identified and analyzed \textit{closed frequent patterns} among all sequences of behavior codes in successful and unsuccessful communication sequences. A \textit{frequent pattern} is \textit{closed} if none of its supersets have the same support count~\cite{pasquier1999discovering}, where a set X is a superset of another set Y, if X contains all the elements of the set Y. For example, the itemsets \{A\}, \{C\}, \{D\} and \{C, D\} in Figure~\ref{fig:fpm} are not closed frequent patterns since their supersets \{A, B\}, \{B, C\}, \{B, D\} and \{B, C, D\} have the same support count. Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\} and \{B, C, D\} are closed frequent patterns since none of their supersets have the same support count. On the other hand, itemsets \{A, C\}, \{A, B, C\}, \{A, B, D\}, \{A, C, D\} and \{A, B, C, D\} have support counts of 1, 0, 1, 0 and 1, respectively, which is less than the minimum support count and thus are identified as non-frequent itemsets. Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed prior work~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences. For each pattern, the statistical significance of the difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value of Pearson's chi-square test.
